---
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, warning = FALSE, message = FALSE)
```


## Greenhouse gas emissions and combustible fuel electricity production in New Zealand and other OECD countries

### Disclaimer
This blog post was from a written exercise from my previous job application. This post aims to demonstrate the process of data manipulation, analysis, and visualisation using the R programming language as the underlying data is considered outdated.  

### Executive Summary

+ Chart shows a strong positive relationship between annual combustible fuel electricity production and annual greenhouse gas emissions across 30 OECD countries.
+ Comparing across 30 OECD countries, New Zealand (NZ) ranked 6th-lowest in greenhouse gas emissions and 11th-lowest in combustible fuel electricity production (per 1000 population per annum).

This analysis draws on three data sources: (1) OECD annual population statistics, (2) International Energy Agency monthly electricity statistics, and (3) Energy Industries yearly greenhouse gas emissions **between 2010 and 2012**. This report describes the link between average annual greenhouse gas emissions (in gigagram; Gg) and combustible fuel electricity production (in gigawatts hour; GWh) per 1000 population across OECD countries, and then compares NZ rates with rates from other OECD countries.

Across OECD countries, the analysis identifies extremely strong evidence that as annual combustible fuel electricity production increases; the amount of annual greenhouse gas emissions also increases. We estimate that for every one GWh increase in the former, the latter increases by 0.876 Gg (per 1000 population).

The chart shows the annual average levels of the greenhouse gas emissions versus combustible fuel electricity production for NZ and other OECD countries. NZ generates greenhouse gas emissions of 1.58 Gg per 1000 population each year (shown by blue horizontal dashed line) and combustible fuel electricity production of 2.63 GWh per 1000 population each year (shown by blue vertical dashed line), which are both significantly lower than countries like Australia, the United States, and the United Kingdom. On both measures, only Iceland, Switzerland and France ranked significantly better. NZ appears to produce significantly more power (per 1000 population) using combustible fuels than Norway, Slovakia and Hungary, but generates less greenhouse gas emissions (per 1000 population) than these countries.



### Load the required packages

```{r load_library}
library(countrycode)
library(broom)
library(knitr)
library(ggrepel)
library(tidyverse)
```


# Extract datasets 

These datasets were converted to CSV for importing.

```{r pressure}

pop_raw <- read_csv("data/DP_LIVE_01062019092109456.csv")
Carbon_Gg_emissions_raw <- read_csv("data/GHGs from Energy Industries.csv")
power_usage_raw <- read_csv("data/Monthly Electricity Statistics.csv")
```


## Clean-up the power_usage dataset 

```{r}
power_usage_by_country_year <- 
  power_usage_raw %>% 
  separate(Index, c("Country", "Type", "Fuel_Type"), sep = "\\." ) %>% 
  filter(Fuel_Type == "THERMAL") %>% 
  select(-Type) %>% 
  gather(key = month_year, val = Energy_GWh, -Country, -Fuel_Type ) %>% 
  separate(month_year, c("Month", "Year"), sep = "-" ) %>% 
  group_by(Country, Fuel_Type, Year) %>%
  summarise(Energy_GWh = sum(Energy_GWh, na.rm = TRUE)) %>% 
  ungroup() %>% 
  mutate(Country = if_else(Country == "NETHLAND", "NETHERLA",  
                           if_else(Country == "KOREA", "SOUTH KO", Country)))
```

## Clean-up the Carbon_Gg_emissions_raw dataset   

```{r}
Carbon_Gg_emissions_by_country_year <- 
  Carbon_Gg_emissions_raw %>% 
  select(Country, `2010`:`2012`) %>% 
  gather(key = Year, val = Carbon_Gg, -Country)  %>% 
  mutate(
    Country_label = Country,
    Country = toupper(str_sub(Country, 1, 8)),
    Year = str_sub(Year, 3, 4)) %>% 
  mutate(Country = 
           if_else(Country == "CZECH RE", "CZECH",
                   if_else(Country == "NEW ZEAL", "NZ",  
                           if_else(Country == "UNITED K", "UK", 
                                   if_else(Country == "UNITED S", "USA",
                                           if_else(Country == "SWITZERL",
                                                   "SWITLAND", Country)))))) 
```

## Clean-up the population dataset

```{r}
pop_by_country_year <- 
  pop_raw %>% 
  select(LOCATION, TIME, Value) %>% 
  rename(Country = LOCATION, 
         Year = TIME) %>% 
  mutate(Country = countrycode(Country, "iso3c", "country.name", warn = FALSE),
         Pop = Value * 1e6 ) %>% 
  mutate(Country = toupper(str_sub(Country, 1, 8)),
         Year = str_sub(Year, 3, 4)) %>% 
  mutate(Country = 
           if_else(Country == "CZECH RE", "CZECH",
                   if_else(Country == "NEW ZEAL", "NZ",  
                           if_else(Country == "UNITED K", "UK", 
                                   if_else(Country == "UNITED S", "USA",
                                           if_else(Country == "SWITZERL", "SWITLAND",
                                                   if_else(Country == "CZECHIA", 
                                                           "CZECH", Country)))))))
```


## Combined the three datasets by country name 

And standardised the measurements to per 1000 person. 
The country code are joined based on the power usage dataset, which is the first 8 upper case letter of the country name.

```{r}
dat_combined <- 
  power_usage_by_country_year %>% 
  inner_join(Carbon_Gg_emissions_by_country_year) %>% 
  inner_join(pop_by_country_year) %>% 
  mutate(Energy_GWh_perC = Energy_GWh/Pop * 1000,
         Carbon_Gg_perC = Carbon_Gg/Pop * 1000)

```

This dataframe contains the non-matching country codes, so I can manually correct the names of countries for merging. 

```{r}
check_nonmatching_countries <-
  power_usage_by_country_year %>%
  select(Country, Energy_GWh, Year) %>%
  full_join(Carbon_Gg_emissions_by_country_year %>%
              select(Country, Carbon_Gg, Year)) %>%
  full_join(pop_by_country_year %>%
              select(Country, Pop, Year)) %>% 
  mutate(Energy_GWh = is.na(Energy_GWh),
         Carbon_Gg = is.na( Carbon_Gg),
         Pop = is.na(Pop)) %>% 
  distinct(Country, Energy_GWh, Carbon_Gg, Pop) %>% 
  filter(Energy_GWh | Carbon_Gg|Pop)
```

## Linear regression to estimate the trend

Modeling for combustible fuel electricity production  and the amount of greenhouse gas emissions 

```{r}
fit <- lm(Carbon_Gg_perC ~ Energy_GWh_perC, dat_combined )
summary(fit)
```
The analysis identifies extremely strong evidence that  as annual combustible fuel electricity production increases by 1 GWh, The amount of annual greenhouse gas emissions also increases by 0.87577 Gg.

## Figure Electricity production using combustible fuels versus Greenhouse emissions

```{r}
ggplot(dat_combined, aes(x = Energy_GWh_perC, y = Carbon_Gg_perC)) +
  geom_point() +
  labs(x = "Annual electricity production (GWh) per 1,000 people using combustible fuels",
       y = "Annual greenhouse emissions (Gg) per 1,000 people") +
  geom_smooth(method = "lm") +
  theme_bw()
```

Decided not to use this plot for the presentaion, even though it shows a positive linear trend but it does not help in comparing NZ against other countries.


## One way ANOVA comparing against NZ 

### Set NZ as reference group for comparison 

```{r}
country_list <- unique(dat_combined$Country_label)

dat_combined$Country_label <- 
  factor(dat_combined$Country_label, 
         levels = c("New Zealand" ,
                    country_list[country_list != "New Zealand" ]))
```

### Fitting models to compare against NZ

The average Greenhouse gas emissions and  combustible fuel electricity production of NZ against other OECD countries per year

```{r}
fit_Carbon_Gg <- lm(Carbon_Gg_perC ~ Country_label,
                    dat_combined)
```

### Difference of average Greenhouse gas emissions of other countries against NZ

```{r}
tidy(fit_Carbon_Gg) %>% 
  arrange(estimate) %>% 
  knitr::kable("markdown")
```

NZ is significantly worse than Iceland, Switzerland and France in average Greenhouse gas emissions.

```{r}
fit_Energy_GWh <- lm(Energy_GWh_perC ~  Country_label,
                     dat_combined )
```

### Difference of average combustible fuel electricity production of other countries against NZ

```{r}
tidy(fit_Energy_GWh) %>% 
  arrange(estimate) %>% 
  knitr::kable("markdown")
```

NZ is significantly worse than Iceland, Switzerland, France, Norway, Slovakia, and Hungary in average combustible fuel electricity production.


## Fitting models

To estimate the average Greenhouse gas emissions and combustible fuel electricity production  for different countries per year for the chart. 

```{r}
fit_Carbon_Gg <- lm(Carbon_Gg_perC ~ 0 + Country_label, dat_combined)

fit_Energy_GWh <- lm(Energy_GWh_perC ~  0 + Country_label, dat_combined )
```




## Barchart comparing between countries 

Extracting the average annual Greenhouse gas emissions and combustible fuel electricity production for each country.  Then stacking the results of annual Greenhouse gas emissions and combustible fuel electricity production for plotting 

```{r}
dat_plot <- 
  list("Greenhouse gas emissions" = 
         fit_Carbon_Gg,
       "Combustible fuels electricity production" = 
         fit_Energy_GWh) %>% 
  map(function(x) tidy(x) %>% 
        filter(!term %in% c("(Intercept)"))%>% 
        mutate(term = str_replace(term, "Country_label", ""))  %>%
        mutate(Country = fct_reorder(term, estimate, .desc = TRUE))) %>% 
  bind_rows(.id = "Type") 
```



### Reference lines for average measures for NZ

```{r}
ref_line <- 
  list( " " = tidy(fit_Energy_GWh)[1,2],
        "NZ annual average" = tidy(fit_Carbon_Gg)[1,2]) %>% 
  bind_rows(.id = "Type") 
```
### Side-by-side bar-plot comparing the annual Greenhouse gas emissions and combustible fuel electricity production between 30 OECD countries

```{r, fig.height = 10, fig.width = 8}
g <- 
  ggplot(dat_plot, aes(x =  Country, y = estimate, fill = Type)) + 
  geom_col(position = "dodge") + 
  geom_hline(aes(yintercept = estimate, col = Type), data = ref_line,
             linetype = "dashed", size = 1) +
  labs(title = "Greenhouse gas emissions and combustible fuel electricity production \n in New Zealand and other OECD countries",
    y = "Average annual amount of greenhouse gas emissions (in Gg) or \n combustible fuel electricity production (in GWh) (per 1000 population)",
    fill = "", col = "") + 
  coord_flip() +  
  #theme_bw()+ 
  theme(legend.position = "top",
        axis.text.y =
          element_text(
            face =ifelse(levels(dat_plot$Country) ==
                           "New Zealand", "bold", "plain"))) 

g
```

Decided not to use it, because the association between the average annual Greenhouse gas emissions and combustible fuels electricity production is not obvious in this plot. 


## Scatterplot comparing between countries 

Scatterplot shows the annual average levels of the greenhouse gas emissions versus combustible fuel electricity production for NZ and other OECD countries.

Re-extracting the average annual Greenhouse gas emissions and combustible fuel electricity production for each country for the final chart

```{r}
Carbon_Gg_average <- 
  tidy(fit_Carbon_Gg) %>% 
  select(term, estimate) %>% 
  filter(!term %in% c("(Intercept)"))%>% 
  mutate(term = str_replace(term, "Country_label", "") )%>% 
  rename(Country = term, 
         Carbon_Gg = estimate)


Carbon_Gg_average %>% 
  arrange(Carbon_Gg) %>% 
  print(n=Inf)
```
NZ ranked 6th with 1.58 Gg


```{r}
Energy_GWh_average <- 
  tidy(fit_Energy_GWh)  %>% 
  select(term, estimate) %>% 
  filter(!term %in% c("(Intercept)"))%>% 
  mutate(term = str_replace(term, "Country_label", "") )%>% 
  rename(Country = term, 
         Energy_GWh = estimate)

Energy_GWh_average%>% 
  arrange(Energy_GWh) %>% 
  print(n=Inf)
```

NZ ranked 11th with 2.63 GWh

```{r}
dat_plot <- 
  Energy_GWh_average %>%
  full_join(Carbon_Gg_average) %>% 
  mutate(Country = factor(Country))
```


### Final product

```{r, fig.height= 7, fig.width=10}
g <- 
  ggplot(dat_plot, aes(x =  Energy_GWh, y = Carbon_Gg, 
                       label = Country)) + 
  geom_point(size =  ifelse(dat_plot$Country == "New Zealand", 3, 1)) + 
  geom_text_repel(
    size =  ifelse(dat_plot$Country == "New Zealand", 5, 4),
    fontface = ifelse(dat_plot$Country == "New Zealand", "bold", "plain")) +
  labs(title = 
         "Average annual combustible fuel electricity production versus greenhouse gas emissions for each OECD country",
       x = "Average annual combustible fuel electricity production (GWh) per 1,000 people",
       y = "Average annual amount of greenhouse gas emissions (Gg) per 1,000 people") +
  geom_smooth(method = "lm", se = FALSE, size = .5) +
  geom_hline(yintercept = as.numeric(dat_plot[1,3]),
             linetype = "dashed", col = "steelblue") +
  geom_vline(xintercept = as.numeric(dat_plot[1,2]),
             linetype = "dashed", col = "steelblue") + 
  annotate("rect", xmin = -0.5, xmax = as.numeric(dat_plot[1,2]),  
                    ymin = -0.5, ymax = as.numeric(dat_plot[1,3]),  
           fill = "steelblue", alpha = 0.1)+
  annotate("rect", xmin = as.numeric(dat_plot[1,2]) , xmax = 9.5,  
           ymin = as.numeric(dat_plot[1,3]), ymax = 11,  
           fill = "steelblue", alpha = 0.1)+
  annotate("text", x = 1.55, y = 8, size = 4, col = "steelblue", fontface = "bold",
           label = "Less power production,\n but more emissions than NZ.") +
  annotate("text", x = 1.55, y = 0, size = 4, col = "steelblue",  fontface = "bold",
           label = "Less power production,\n and less emissions than NZ.") +
  annotate("text", x = 5, y = 8, size = 4, col = "steelblue",  fontface = "bold",
           label = "More power production,\n and more emissions than NZ.") +
  annotate("text", x = 5, y = 0, size = 4, col = "steelblue",  fontface = "bold",
           label = "More power production,\n but less emissions than NZ.") +
  scale_x_continuous(breaks = seq(0, 9.5, by = 1), expand = c(0,0), limits = c(-0.5, 9.5)) + 
  scale_y_continuous(breaks = seq(0, 11, by = 1), expand = c(0,0), limits = c(-0.5, 11)) + 
  theme_bw()

g
```

### Notes:
 + The point and Font size for NZ have been made bigger.
 + geom_text_repel function is to avoid overlapping of the country names.
 + Smoother with linear regression line to show the positive linear association. 
 + Horizontal and vertical lines allow the reader to compare against measures for NZ.
 + Annotate function allows me to have shading and write text on the plot in specific area.
 + The continuous scales for both x and y axis are adjusted to 1 unit break.
 + Lastly, I used black and white theme, so it is easier to read.
 
 
## Citation
OECD (2021), Population (indicator). doi: 10.1787/d434f82b-en (Accessed on 12 December 2021)
